Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-30 survey/survey.tex
smoothen transition in sec 3

What does it mean for two finite sets to be ``the same''?  Well,
clearly the sets $A = \left\{1, 2, 3\right\}$ and
$B = \left\{c, d\right\}$ are different.  Why?  Well, supposed there
was a transformation $f : A \rightarrow B$ that deformed $A$ into $B$,
and another $g : B \rightarrow A$ which undid this
transformation. Since $f$ is total, by the pigeonhole principle, two
elements of $A$ would be mapped to the same element of $B$. Suppose
that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
cannot be both $2$ and $3$, and so $g$ is not the inverse of $f$. With
just a little more work, we can show that $f$ (and $g$) must be both
injective and surjective. In other words, $f$ (and $g$) must be a
bijection between $A$ and $B$.  And of course this only happens when
$A$ and $B$ have the same number of elements. More importantly, given
a bijection $f : C \rightarrow D$ of finite sets $C,D$, there always
exists another bijection $g : D \rightarrow C$ which is $f$'s
inverse. So, for finite sets, \emph{bijections} act as reversible
deformations.

@@

In programming practice we are however more interested in structured
data which will be the subject on next section. What we learn from
this short investigation using untyped flat sets is what the
\emph{purely operational} view of the theory would be. In particular,
it tells us that permutations are an inescapable part of the fabric of
reversible computing.  However as permutations are untyped, and act on
the canonicalized version of $n$-element sets (i.e. those sets where
all the structure has been forgotten), these are a rather pale shadow
of the rich tapestry of information-perserving transformations of
structured data which we investigate next.

@@

Instead of the spaces consisting solely of unstructured isolated
points, we now investigate structured spaces built from sums and
products of elementary spaces. This structure corresponds to the
building blocks of type theory which are: the empty type
($\mathbf{0}$), the unit type ($\mathbf{1}$), the sum type ($\uplus$),
and the product ($\times$) type. Before getting into the formal
theory, let's consider the space
$(\mathbf{1} \uplus \mathbf{0}) \times (\mathbf{1} \uplus
\mathbf{1})$. The subspace $(\mathbf{1} \uplus \mathbf{0})$ is the sum
of the space $\mathbf{1}$ containing one element and the empty space
$\mathbf{0}$. By construction this space is different from just the
space $\mathbf{1}$ but one can argue that the spaces are equivalent in
some sense.

\amr{new wavefront for new pass}

These have a nice correspondance (Curry-Howard) to logic:

>>>>

What does it mean to deform a space of points? For example, what
transformation can we do on a bag or marbles? Well, we can shuffle
them around and that is the only transformation that will preserve the
space. Turning to the mathematical abstraction as sets, we ask what
does it mean for two finite sets to be ``the same''?  Well, clearly
the sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$
are different.  Why?  Well, supposed there was a transformation
$f : A \rightarrow B$ that deformed $A$ into $B$, and another
$g : B \rightarrow A$ which undid this transformation. Since $f$ is
total, by the pigeonhole principle, two elements of $A$ would be
mapped to the same element of $B$. Suppose that this is $2$ and $3$,
and that they both map to $d$.  But $g(d)$ cannot be both $2$ and $3$,
and so $g$ is not the inverse of $f$. With just a little more work, we
can show that $f$ (and $g$) must be both injective and surjective. In
other words, $f$ (and $g$) must be a bijection between $A$ and $B$.
And of course this only happens when $A$ and $B$ have the same number
of elements. More importantly, given a bijection $f : C \rightarrow D$
of finite sets $C,D$, there always exists another bijection
$g : D \rightarrow C$ which is $f$'s inverse. So, for finite sets,
\emph{bijections} act as reversible deformations.

@@

In programming practice we are however more interested in structured
data which will be the subject on next section. What we learn from
this short investigation using untyped flat sets is what the

'\emph{purely operational} view of the theory would be. In particular,
it tells us that permutations are an inescapable part of the fabric of
reversible computing.  However as permutations are untyped, and act on
the canonicalized version of $n$-element sets (i.e. those sets where
all the structure has been forgotten), these are a rather pale shadow
of the rich tapestry of information-perserving transformations of
structured data which we investigate next.

@@

Instead of the spaces consisting solely of unstructured isolated
points, we now investigate structured spaces built from sums and
products of elementary spaces. This structure corresponds to the
building blocks of type theory which are: the empty type
($\mathbf{0}$), the unit type ($\mathbf{1}$), the sum type ($\uplus$),
and the product ($\times$) type. Before getting into the formal
theory, let's consider possible deformations on the space
$(\mathbf{1} \uplus \mathbf{0}) \times (\mathbf{1} \uplus
\mathbf{1})$. This space is the product of two subspaces: the subspace
$(\mathbf{1} \uplus \mathbf{0})$ which itself is the sum of the space
$\mathbf{1}$ containing one element and the empty space $\mathbf{0}$
and the subspace $(\mathbf{1} \uplus \mathbf{1})$ which is the sum of
two spaces each containing one element. First, as discussed in the
previous section, any deformation of this space must at least preserve
the number of elements: we can neither create nor destroy points
during any continuous deformation. Seeing that the number of elements
in our example space is 2, a reasonable hypothesis is that we can
deform the space above to any other space with 2 elements. What this
really means is that we are treating the sum and product structure as
malleable. For example, imagining a product structure as arranged in a
grid, by stretching we can turn this structure to a sum structure
arranged in a line.

We now turn into the mathematical formalization of this idea. First we
note that the type structure of types has a nice correspondance
(Curry-Howard) to logic:
